---
title: "The 'odemodeling' R package"
author: Juho Timonen
date: 21 Feb 2022
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{The 'odemodeling' R package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates core functionality of the package.

```{r setup}
library(odemodeling)
library(ggplot2)
```

# Creating a model

All models need to involve an ODE system of the form
\begin{equation}
    \label{eq: ode}
    \frac{\text{d} \textbf{y}(t)}{\text{d}t} = f_{\psi}\left(\textbf{y}(t), t\right),
\end{equation}
where $f_{\psi}: \mathbb{R}^D \rightarrow \mathbb{R}^D$ with parameters $\psi$. As an example we define an ODE system 
\begin{equation}
    \label{eq: sho}
    f_{\psi}\left(\textbf{y}, t\right) = 
    \begin{bmatrix}
      y_2 \\
      - y_1 - \theta y_2
    \end{bmatrix}
\end{equation}
describing a simple harmonic oscillator, where $\psi = \{ k \}$ and dimension $D = 2$. The Stan code for the body of this function is
```{r sho_body}
sho_fun_body <- "
  vector[2] dy_dt;
  dy_dt[1] = y[2];
  dy_dt[2] = - y[1] - k*y[2];
  return(dy_dt);
"
```

We need to
define the variable for the initial system state  at `t0` as `y0`. The ODE
system dimension is declared as `D` and number of time points as `N`.
```{r sho_dims}
N <- stan_dim("N", lower = 1)
D <- stan_dim("D")
y0 <- stan_vector("y0", length = D)
k <- stan_param(stan_var("k", lower = 0), "inv_gamma(5, 1)")
```

Finally we declare the parameter `k` and its prior.
```{r sho_param}
k <- stan_param(stan_var("k", lower = 0), prior = "inv_gamma(5, 1)")
```

The following code creates and compiles the Stan model.
```{r sho_create_prior}
sho <- ode_model(N,
  odefun_vars = list(k),
  odefun_body = sho_fun_body,
  odefun_init = y0
)
print(sho)
```
As we see, all variables that affect the function $f_{\psi}$ need to be given
as the `odefun_vars` argument. The function body itself is then the `odefun_body` argument. In this function body, we can use the following variables without having to declare them or write Stan code that computes them:

* The ODE state `y`, which is a vector of same length as dimension of `y0`.
* Any variables that we give as `odefun_vars` for `ode_model`.
* Any variables that are dimensions of `odefun_vars`.

The initial state needs to be given as `odefun_init`. See the documentation of the `ode_model` function for more information.

We could call `print(sho$stanmodel)` to see the entire generated Stan model code.

# Sampling from prior

We can sample from the prior distribution of model parameters like so.

```{r sho_prior_sample}
sho_fit_prior <- sho$sample(
  t0 = 0.0,
  t = seq(0.1, 10, by = 0.1),
  data = list(y0 = c(1, 0), D = 2),
  refresh = 0,
  solver = rk45(
    abs_tol = 1e-13,
    rel_tol = 1e-13,
    max_num_steps = 1e9
  )
)
```

We can view a summary of results 
```{r sho_fit_print}
print(sho_fit_prior)
```

We can obtain the ODE solution using each parameter draw by doing
```{r sho_fit_view_ode1}
ys <- sho_fit_prior$extract_odesol_df()
```

We can plot ODE solutions like so
```{r sho_fit_view_ode2, fig.width=6.5, fig.height=3.2}
sho_fit_prior$plot_odesol(alpha = 0.3)
```

We can plot the distribution of ODE solutions like so
```{r sho_fit_view_ode3, fig.width=6.5, fig.height=3.2}
sho_fit_prior$plot_odesol_dist(include_y0 = TRUE)
```

We can plot ODE solution using one draw like so
```{r sho_fit_view_ode4, fig.width=6.5, fig.height=3.2}
sho_fit_prior$plot_odesol(draw_inds = 45)
```

# Using different ODE solvers
We generate quantities using a different `solver` and different output time points `t`. Possible solvers are `rk45()`,`bdf()`, `adams()`, `ckrk()`, `midpoint()`, and
`rk4()`. Of these the first four are adaptive and built-in to Stan, where as
the last two take a fixed number of steps and are written in Stan code.

```{r sho_gq_bdf, fig.width=6.5, fig.height=3.2}
gq_bdf <- sho_fit_prior$gqs(solver = bdf(tol = 1e-4), t = seq(0.5, 10, by = 0.5))
```

```{r sho_gq_mp, fig.width=6.5, fig.height=3.2}
gq_mp <- sho_fit_prior$gqs(solver = midpoint(num_steps = 4), t = seq(0.5, 10, by = 0.5))
```

We can again plot ODE solution using one draw like so
```{r sho_view_gq, fig.width=6.5, fig.height=3.2}
gq_bdf$plot_odesol(draw_inds = 45)
gq_mp$plot_odesol(draw_inds = 45)
```

# Defining a likelihood

Next we assume that we have some data vector `y1_obs` and define a likelihood
function. 
```{r sho_loglik}
sho_loglik_body <- "
  real loglik = 0.0;
  for(n in 1:N) {
    loglik += normal_lpdf(y1_obs[n] | y_sol[n][1], sigma);
  }
  return(loglik);
"
```

In this function body, we can use the following variables without
having to declare them or write Stan code that computes them:

* The ODE solution `y_sol`.
* Any variables that we give as `loglik_vars` for `ode_model`.
* Any variables that are dimensions of `loglik_vars`.

Here we define as `loglik_vars` a `sigma` which is a noise magnitude parameter, and the data `y1_obs`. Notice that we can also use the `N` variable in `loglik_body`, because it is a dimension (length) of `y1_obs`.

```{r sho_loglik_vars}
sigma <- stan_param(stan_var("sigma", lower = 0), prior = "normal(0, 2)")
y1_obs <- stan_vector("y1_obs", length = N)
```

The following code creates and compiles the posterior Stan model.
```{r sho_create}
sho_post <- ode_model(N,
  odefun_vars = list(k),
  odefun_body = sho_fun_body,
  odefun_init = y0,
  loglik_vars = list(sigma, y1_obs),
  loglik_body = sho_loglik_body
)
print(sho_post)
```

# Sampling from posterior

Now if we have some data
```{r sho_data}
y1_obs <- c(
  0.801, 0.391, 0.321, -0.826, -0.234, -0.663, -0.756, -0.717,
  -0.078, -0.083, 0.988, 0.878, 0.300, 0.307, 0.270, -0.464, -0.403,
  -0.295, -0.186, 0.158
)
t_obs <- seq(0.5, 10, by = 0.5)
```
and assume that initial state `y0 = c(1, 0)` is known, we can fit the model
```{r sho_fit_post}
sho_fit_post <- sho_post$sample(
  t0 = 0.0,
  t = t_obs,
  data = list(y0 = c(1, 0), D = 2, y1_obs = y1_obs),
  refresh = 0,
  solver = midpoint(2)
)
```

We fit the posterior distribution of ODE solutions against the data
```{r sho_post_plot, fig.width=6.5, fig.height=3.2}
plt <- sho_fit_post$plot_odesol_dist()
df_data <- data.frame(t_obs, y1_obs, ydim = rep("y1", length(t_obs)))
colnames(df_data) <- c("t", "y", "ydim")
df_data$ydim <- as.factor(df_data$ydim)
plt <- plt + geom_point(data = df_data, aes(x = t, y = y), inherit.aes = FALSE)
plt
```

# Reliability of ODE solver

Finally we can study whether the solver we used during MCMC (`midpoint(2)`) was accurate enough. This is done by solving the system using increasingly more numbers of steps in
the solver, and studying different metrics computed using the ODE solutions and corresponding likelihood values.
```{r sho_rel}
solvers <- midpoint_list(c(4, 6, 8, 10, 12, 14, 16, 18))
rel <- sho_fit_post$reliability(solvers = solvers)
print(rel$metrics)
unlink("results")
```

The `mad_odesol` and `mad_loglik` are the maximum absolute difference in the ODE solutions and log likelihood, respectively, over all MCMC draws. The former is denoted *MAE* in *Timonen et. al (2022)*. Please refer to that paper in order to interpret the `pareto_k` column. Briefly we note that the Pareto-k values seem to be converging to a value smaller than 0.5, meaning that importance sampling is possible and we don't need to run MCMC again.

# References

* Timonen et. al (2022). *An importance sampling approach for reliable and efficient inference in Bayesian ordinary differential equation models*. [arXiv](https://arxiv.org/abs/2205.09059)

