---
title: "Defining and fitting ODE models using the odemodeling package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining and fitting ODE models using the odemodeling package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(odemodeling)
```

## Simple harmonic oscillator

### Creating a model

We define an ODE system with only one parameter `theta`. We need to
define the variable for the initial system state at `t0` as `y0`. The ODE
system dimension is declared as `D` and number of time points as `N`.

```{r sho_define}
N <- stan_dim("N", lower=1)
D <- stan_dim("D")
y0 <- stan_vector("y0", length = D)
theta <- stan_param(stan_var("theta", lower=0), "theta ~ inv_gamma(5, 1);")
sho_fun_body <- "
  vector[2] dy_dt;
  dy_dt[1] = y[2];
  dy_dt[2] = - y[1] - theta*y[2];
  return(dy_dt);
"
```

The following code creates and compiles the Stan model.
```{r sho_create}
sho <- create_odemodel(N,
                       odefun_vars = list(theta), 
                       odefun_body = sho_fun_body,
                       odefun_init = y0)
print(sho)
```

We print the full Stan code of the model.

```{r sho_print}
sho$stanmodel
```

### Sampling from prior

We can sample from the prior distribution of model parameters like so.

```{r sho_prior_sample}
sho_fit_prior <- sample_odemodel(sho, 
                             t0 = 0.0, 
                             t = seq(0.1, 10, by = 0.1),
                             data = list(y0 = c(1, 0), D = 2),
                             refresh = 0,
                             solver_conf = list(
                               abs_tol = 1e-13,
                               rel_tol = 1e-13,
                               max_num_steps = 1e9
                             )
                             )
```

We can view a summary of results 
```{r sho_fit_print}
print(sho_fit_prior)
```

We can obtain the ODE solution using each parameter draw by doing
```{r sho_fit_view_ode}
ys <- sho_fit_prior$extract_odesol()
print(dim(ys))
plot(ys[12,,2])
plot(ys[12,,1])
```


